### 2018-05-15 submission/manuscript.Rmd
complete removal of sub-sampling
fix cycle number jargon

**Background.** Bias is introduced at different stages of the 16S rRNA gene sequencing workflow. Commonly studied sources of bias include preservation and DNA extraction methods. Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow. Here, we examine how both cycle number and HiFi DNA polymerase can change the bacterial community and introduce bias to the final obtained results.

@@

**Conclusions.** HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence makes comparison across studies difficult. Care should be exercised when choosing both HiFi DNA polymerase and cycle number to be used in 16S rRNA gene sequencing studies.

@@

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000]. Many of these sources of biases are made worse as cycle number increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, the selective amplification of AT-rich over GC-rich sequences can exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC [@polz_bias_1998]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will artificially increase community diversity [@Haas2011]. Although these differences are not necessarily dependent on primer and DNA polymerase used, there are also biases that are.

@@

***Analysis Workflow:***
All samples were rarefied to 1000 sequence. The total number of OTUs was analyzed for both the fecal and mock community samples. For fecal samples, cycle dependent affects on Bray-Curtis indices were assessed for cycle group and within individual differences from the previous cycle (e.g. 20 versus 25, 25 versus 30, etc.). Based on these observations we analyzed potential reasons for these differences. Analysis of the mock community of each HiFi DNA polymerase for general sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras were assessed before and after chimera removal. Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed.

@@

There were few differences that depend on HiFi DNA polymerase for between sample diversity, as measured by the Bray-Curtis index. Using this metric our observations found no differences in the overall bacterial community composition for sub-sampling depth or cycle number used. One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group. Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases. When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. With differences in the number of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it might be easier to avoid specific DNA polymerase families altogether.

>>>>

**Background.** Bias is introduced at different stages of the 16S rRNA gene sequencing workflow. Commonly studied sources of bias include preservation and DNA extraction methods. Although the number of cycles and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow. Here, we examine how both cycle number and HiFi DNA polymerase can change the bacterial community and introduce bias to the final obtained results.

@@

**Conclusions.** HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence makes comparison across studies difficult. Care should be exercised when choosing both HiFi DNA polymerase and number of cycles to be used in 16S rRNA gene sequencing studies.

@@

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000]. Many of these sources of biases are made worse as the number of cycles increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, the selective amplification of AT-rich over GC-rich sequences can exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC [@polz_bias_1998]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as the number of cycles increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will artificially increase community diversity [@Haas2011]. Although these differences are not necessarily dependent on primer and DNA polymerase used, there are also biases that are.

@@

***Analysis Workflow:***
All samples were rarefied to 1000 sequence. The total number of OTUs was analyzed for both the fecal and mock community samples. For fecal samples, cycle dependent affects on Bray-Curtis indices were assessed for cycle group and within individual differences from the previous cycle (e.g. 20 versus 25, 25 versus 30). Based on these observations we analyzed potential reasons for these differences. Analysis of the mock community of each HiFi DNA polymerase for general sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras were assessed before and after chimera removal. Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed.

@@

There were few differences that depend on HiFi DNA polymerase for between sample diversity, as measured by the Bray-Curtis index. Using this metric our observations generally found no differences in the overall bacterial community composition for the number of cycles used. One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group. Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases. When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. With differences in the number of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it might be easier to avoid specific DNA polymerase families altogether.

### 2018-05-15 submission/manuscript.tex
fix cycle number jargon

\textbf{Background.} Bias is introduced at different stages of the 16S
rRNA gene sequencing workflow. Commonly studied sources of bias include
preservation and DNA extraction methods. Although cycle number and high
fidelity (HiFi) DNA polymerase are studied less often, they are still
important sources of bias in this workflow. Here, we examine how both
cycle number and HiFi DNA polymerase can change the bacterial community
and introduce bias to the final obtained results.

@@

\textbf{Conclusions.} HiFi DNA polymerase dependent differences in the
number of OTUs and chimera prevalence makes comparison across studies
difficult. Care should be exercised when choosing both HiFi DNA
polymerase and cycle number to be used in 16S rRNA gene sequencing
studies.

@@

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results (Eckert
\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
made worse as cycle number increases (Wang \& Wang, 1996; Haas et al.,
2011; Kebschull \& Zador, 2015). For example, the selective
amplification of AT-rich over GC-rich sequences can exaggerate the
difference between 16S rRNA genes higher in AT versus those higher in GC
(Polz \& Cavanaugh, 1998). Both amplification error and non-specific
amplification (e.g.~incorrect amplicon size products) can also increase
as cycle number increases which can drastically change commonly used
diversity measures (Acinas et al., 2005; Santos et al., 2016).
Additionally, chimeras can form from an aborted extension step followed
by a subsequent priming error and secondary extension and will
artificially increase community diversity (Haas et al., 2011). Although
these differences are not necessarily dependent on primer and DNA
polymerase used, there are also biases that are.

>>>>

\textbf{Background.} Bias is introduced at different stages of the 16S
rRNA gene sequencing workflow. Commonly studied sources of bias include
preservation and DNA extraction methods. Although the number of cycles
and high fidelity (HiFi) DNA polymerase are studied less often, they are
still important sources of bias in this workflow. Here, we examine how
both cycle number and HiFi DNA polymerase can change the bacterial
community and introduce bias to the final obtained results.

@@

\textbf{Conclusions.} HiFi DNA polymerase dependent differences in the
number of OTUs and chimera prevalence makes comparison across studies
difficult. Care should be exercised when choosing both HiFi DNA
polymerase and number of cycles to be used in 16S rRNA gene sequencing
studies.

@@

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results (Eckert
\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
made worse as the number of cycles increases (Wang \& Wang, 1996; Haas
et al., 2011; Kebschull \& Zador, 2015). For example, the selective
amplification of AT-rich over GC-rich sequences can exaggerate the
difference between 16S rRNA genes higher in AT versus those higher in GC
(Polz \& Cavanaugh, 1998). Both amplification error and non-specific
amplification (e.g.~incorrect amplicon size products) can also increase
as the number of cycles increases which can drastically change commonly
used diversity measures (Acinas et al., 2005; Santos et al., 2016).
Additionally, chimeras can form from an aborted extension step followed
by a subsequent priming error and secondary extension and will
artificially increase community diversity (Haas et al., 2011). Although
these differences are not necessarily dependent on primer and DNA
polymerase used, there are also biases that are.
